---
title: "csDAM: Cosine Similarity-based Discriminative Analysis of Mixtures"
author: "Dongping Du"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{csDAM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# Overview

Reference-free deconvolution estimates cell-type proportions and cell-type-specific expression profiles from bulk omic data without predefined reference signatures. This vignette introduces **csDAM** (*Cosine Similarity-based Discriminative Analysis of Mixtures*), an unsupervised framework designed for highly mixed samples and weakly separable latent sources.

csDAM follows a directional, cosine-based strategy. Instead of directly factorizing the bulk expression matrix, it first identifies genes whose expression patterns are dominated by a single latent component, then groups genes that vary coherently across samples, and finally selects well-separated candidate directions to estimate proportions using non-negative least squares (NNLS).

This vignette covers:

- the data structure expected by csDAM,
- the main algorithmic ideas,
- a complete end-to-end workflow,
- automatic estimation of the number of latent cell types,
- practical notes for using the package output.

# Getting started

Load the package and inspect the available functions.

```{r}
# library(csDAM)
# ls("package:csDAM")
```

If you are reading this vignette inside the package source tree, the examples below assume that the package is already installed and available on the search path.

# Data structure

csDAM expects a numeric expression matrix with:

- rows = genes or proteins,
- columns = samples.

The matrix should already be normalized and aligned across samples. Non-negative values are strongly preferred because the downstream steps are defined for bulk expression profiles.

```{r}
set.seed(1)

n_genes <- 500
n_samples <- 20
K_true <- 4

# Simulated non-negative example expression matrix
A_true <- matrix(runif(n_samples * K_true, min = 0, max = 1),
                 nrow = n_samples, ncol = K_true)
A_true <- A_true / rowSums(A_true)

S_true <- matrix(rexp(n_genes * K_true, rate = 1),
                 nrow = K_true, ncol = n_genes)

# samples x genes
expr <- A_true %*% S_true
expr <- expr + matrix(rnorm(length(expr), sd = 0.02), nrow = nrow(expr))
expr[expr < 0] <- 0

# genes x samples, which is what csDAM expects
expr <- t(expr)

rownames(expr) <- paste0("gene", seq_len(n_genes))
colnames(expr) <- paste0("sample", seq_len(n_samples))

dim(expr)
expr[1:5, 1:5]
```

# Method summary

csDAM is built around the idea that genes driven by the same latent cell type vary coherently across samples. When a gene is strongly associated with one latent source, its expression vector across samples becomes concentrated in a subset of samples and aligns with the corresponding direction of that source.

At a high level, csDAM performs the following steps:

1. **Sorting**: for each gene, sort expression values in descending order across samples.
2. **Comparing**: compare the sorted expression profile with a family of binary reference vectors using cosine similarity.
3. **Assigning**: assign each gene to the reference support size that maximizes cosine similarity.
4. **Filtering**: remove genes with diffuse patterns that are unlikely to be cell-type-specific.
5. **Grouping**: group genes with similar dominant-sample rankings.
6. **Selecting directions**: choose well-separated candidate directions in angular space.
7. **Rescaling**: estimate proportions with NNLS and normalize the selected directions.

This cosine-based design is robust when latent components are highly mixed or when classical separability assumptions are weak.

# Step 1. Cosine profile curves

The first csDAM step computes, for each gene, a cosine profile over reference vectors of the form $(1, \dots, 1, 0, \dots, 0)$ with varying support size $k$.

```{r}
# cos_mat <- cos_iDEG_fast(
#   expr,
#   transform = "mean_power",
#   gamma = 2
# )
```

The result is a matrix whose rows are genes and whose columns are the support sizes. A larger cosine value indicates a closer match to a more concentrated sample pattern.

## Why cosine similarity?

The manuscript and supplementary theory show that the cosine score is monotonically related to latent gene-level cell-type specificity under the model assumptions used by csDAM. Intuitively, genes that are more specific to one latent source tend to have expression concentrated in fewer samples, which produces a sharper cosine profile.

# Step 2. Identify candidate cell-type-specific genes

After computing cosine profiles, each gene is assigned to the best matching support size.

```{r}
# res_k <- select_genes_by_k(expr, cos_mat)
# names(res_k)
```

The returned object typically contains:

- `selected_genes`: genes retained as candidates,
- `best_k`: the optimal support size for each selected gene,
- `pi`: the permutation that orders samples by decreasing expression for each gene.

Genes with large optimal support sizes are generally diffuse across samples and are removed from later stages.

## Practical note

In the R application code, the default workflow uses the cosine profile matrix together with `best_k` and `pi` to propagate information into grouping and source selection.

# Step 3. Group genes by dominant samples

Genes with similar sample rankings are grouped together, because they are likely driven by the same latent direction.

```{r}
# groups <- group_genes_by_pi(
#   cos_mat,
#   best_k = res_k$best_k,
#   pi_list = res_k$pi,
#   top_n = 5
# )
```

`top_n` controls how many top-ranked samples are used when defining each gene group. Increasing `top_n` makes grouping stricter; decreasing it makes grouping more permissive.

Groups with too few genes are usually discarded, because they are less stable and less informative for downstream source selection.

# Step 4. Select representative directions

The grouped genes usually define more candidate directions than the true number of cell types. csDAM therefore selects a subset of well-separated directions using cosine distance.

```{r}
K <- 4

# sel <- select_sources(
#   groups,
#   expr,
#   method = "cosine",
#   mode = "number",
#   n_sources = K,
#   centroid_genes = 1,
#   per = 1
# )
```

## Selection strategy

The package supports several selection modes, including cosine-based and Spearman-based distances. In the main workflow, cosine distance is the default choice because csDAM is explicitly built around angular separation.

A useful working rule is:

- use `method = "cosine"` for the standard workflow,
- set `per = 1` for proportion estimation,
- use `centroid_genes = 1` when you want each group to be represented by a single marker gene.

## Why angular distance?

csDAM focuses on direction rather than magnitude. After the row-sum normalization step, absolute scale is no longer the main object of interest, so cosine distance is a natural way to identify distinct latent sources.

# Step 5. Estimate proportions

Once representative marker genes or directions are selected, csDAM rescales them into a proportion matrix using NNLS.

```{r}
# prop <- proportion_rescaling(expr, sel)
# A_hat <- prop$A_csCAM
# dim(A_hat)
```

The returned proportion matrix has:

- rows = samples,
- columns = estimated latent cell types.

Depending on the data type and preprocessing, the function name currently returns `A_csCAM` in the application code. In the package documentation and vignette, you can rename the object downstream if you prefer a fully consistent `csDAM` naming scheme.

# Automatic estimation of the number of cell types

csDAM can also estimate the latent dimension automatically.

First rank candidate sources by their distinctness.

```{r}
# ranks <- rank_sources(
#   groups,
#   expr,
#   method = "cosine",
#   centroid_genes = 1,
#   max_n = 20
# )
```

Then detect the primary elbow in the score-versus-complexity trade-off.

```{r}
# K_est <- detect_primary_elbow(
#   ranks,
#   gene_weight = 0.1
# )$primary_rank
# K_est
```

The estimated `K_est` can be passed directly to `select_sources()`.

## When to use automatic K estimation

Automatic estimation is especially useful when the number of latent sources is not known in advance, or when you want to compare the data-driven estimate against a biologically expected number of cell types.

# Complete workflow

The full csDAM workflow is shown below.

```{r}
# cos_mat <- cos_iDEG_fast(expr, transform = "mean_power", gamma = 2)
# res_k <- select_genes_by_k(expr, cos_mat)
# groups <- group_genes_by_pi(
#   cos_mat,
#   best_k = res_k$best_k,
#   pi_list = res_k$pi,
#   top_n = 5
# )
# ranks <- rank_sources(groups, expr, method = "cosine", centroid_genes = 1, max_n = 20)
# K_est <- detect_primary_elbow(ranks, gene_weight = 0.1)$primary_rank
# sel <- select_sources(
#   groups,
#   expr,
#   method = "cosine",
#   mode = "number",
#   n_sources = K_est,
#   centroid_genes = 1,
#   per = 1
# )
# prop <- proportion_rescaling(expr, sel)
# A_hat <- prop$A_csCAM
```

# Interpreting the output

The primary output of csDAM is a proportion matrix, but the intermediate objects are also informative:

- the cosine profile matrix captures how strongly each gene aligns with concentrated reference patterns,
- the selected gene groups represent candidate latent directions,
- the ranking of sources can be used to understand how distinct each candidate direction is,
- the final proportion matrix can be used in downstream statistical analyses.

Typical downstream tasks include:

- comparing cell-type composition across conditions,
- associating latent proportions with phenotypes,
- tracking latent programs across disease stages,
- identifying biologically coherent modules in bulk proteomic or transcriptomic data.

# Practical recommendations

## Preprocessing

csDAM assumes the input has been normalized appropriately. In practice, the best preprocessing depends on the data modality:

- for RNA-seq, use an expression measure that is comparable across samples,
- for proteomics, use a matrix that has already been imputed and normalized,
- avoid negative values whenever possible, because the algorithm is designed for non-negative expression profiles.

## Parameter choices

A few reasonable defaults are:

- `transform = "mean_power"`,
- `gamma = 2`,
- `top_n = 5`,
- `method = "cosine"`,
- `per = 1` for proportion estimation.

For noisier data, a slightly larger `gamma` can emphasize front-loaded expression profiles. For cleaner data, a smaller transformation may be sufficient.

## Output validation

If you have external biological annotations, you can validate csDAM output by checking whether inferred latent directions are enriched for expected cell-type markers or pathways.

# Example: organizing results from the application code

The application script in the package source shows the typical analysis pattern:

```{r eval = FALSE}
cos_mat <- cos_iDEG_fast(data, transform = "mean_power", gamma = 2)
res_k <- select_genes_by_k(data, cos_mat)

groups <- group_genes_by_pi(
  cos_mat, best_k = res_k$best_k, pi_list = res_k$pi, top_n = 5
)

sel <- select_sources(
  groups, data, method = "cosine", mode = "number",
  n_sources = K, centroid_genes = 1, per = 1
)

A_csDAM_raw <- proportion_rescaling(X, sel)$A_csCAM

ranks <- rank_sources(groups, data, method = "cosine", max_n = 20, centroid_genes = 1)

K_csDAM <- detect_primary_elbow(ranks, gene_weight = 0.1)$primary_index
```

This vignette follows the same structure, but adds more context around each step so that the package is easier to understand and adapt.

# Session information

```{r}
sessionInfo()
```
